---
title: "Designing Informative Thermal Performance Experiments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing Informative Thermal Performance Experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5
)
```

## Motivation

Thermal performance curves (TPCs) describe how some measure of performance
(growth, fecundity, survival, etc.) changes with temperature. Fitting a TPC
requires choosing which temperatures to measure performance at, and that
choice matters: temperatures clustered in an uninformative part of the curve
(e.g., only near the thermal optimum) can leave the edges of the curve --
the critical thermal minimum and maximum -- poorly estimated, even with a
large sample size.

TPCdesign helps you choose experimental temperatures that are maximally
informative for fitting a Brière-2 TPC, given a rough prior guess at that
curve's shape. This vignette walks through the full workflow:

1. Define a Brière-2 TPC using rough prior guesses at its parameters.
2. Use `design_temps()` to choose informative temperatures for that curve.
3. Simulate data from the true curve at those temperatures with
   `sim_gamma_data()`.
4. Fit the simulated data and check how well the true parameters are
   recovered, compared to a naive, equally-spaced set of temperatures.

```{r setup}
library(TPCdesign)
```

## The Brière-2 TPC

The Brière-2 curve is parameterized by a lower thermal limit (`ctmin`), an
upper thermal limit (`ctmax`), and shape parameters `a` and `b`.
`briere2_tpc()` evaluates the curve, and `briere2_tpc_Topt()` returns its
thermal optimum:

```{r briere2-curve}
true_ctmin <- 8
true_ctmax <- 38
true_a <- 1
true_b <- 0.2

temp_grid <- seq(true_ctmin, true_ctmax, length.out = 200)
perf <- briere2_tpc(temp_grid, true_ctmin, true_ctmax, true_a, true_b)
topt <- briere2_tpc_Topt(true_ctmin, true_ctmax, true_b)

plot(temp_grid, perf, type = "l", lwd = 2, col = "dodgerblue4",
    xlab = "Temperature", ylab = "Performance",
    main = "Brière-2 thermal performance curve")
abline(v = topt, lty = 2, col = "grey50")
```

In practice, you never know these parameters exactly ahead of time -- you
only have rough guesses, which is exactly what `design_temps()` is built to
work with.

## Choosing informative temperatures

`design_temps()` takes your best guess at `ctmin`, `ctmax`, `a`, and `b`,
along with how uncertain you are about each of `ctmin`, `ctmax`, and `b`
(`ctmin_err`, `ctmax_err`, and `logb_err`, respectively), and searches for
the set of temperatures that is expected to be most informative across that
range of plausible curves. Here we ask for 6 temperatures, treating our
guesses above as uncertain by a few degrees:

```{r design-temps}
set.seed(1)

opt_temps <- design_temps(
    n_temps = 6,
    ctmin = true_ctmin, ctmax = true_ctmax, a = true_a, b = true_b,
    ctmin_err = 2, ctmax_err = 1.5, logb_err = 0.26,
    n_draws = 100L, n_starts = 3L, n_filler = 1L
)
opt_temps
```

For comparison, a naive design might just space temperatures evenly across
the same range:

```{r naive-temps}
naive_temps <- round(seq(true_ctmin + 1, true_ctmax - 1, length.out = 6), 2)
naive_temps
```

```{r compare-temps, echo = FALSE}
plot(temp_grid, perf, type = "l", lwd = 2, col = "grey70",
    xlab = "Temperature", ylab = "Performance",
    main = "Optimized vs. naive temperatures")
points(opt_temps, briere2_tpc(opt_temps, true_ctmin, true_ctmax, true_a, true_b),
      pch = 19, col = "dodgerblue4")
points(naive_temps, briere2_tpc(naive_temps, true_ctmin, true_ctmax, true_a, true_b),
      pch = 17, col = "firebrick")
legend("topleft", legend = c("optimized", "naive"), pch = c(19, 17),
      col = c("dodgerblue4", "firebrick"), bty = "n")
```

Note how the optimized design concentrates more temperatures near `ctmax`,
where the curve is changing quickly and is therefore more informative about
where the curve actually crosses zero, while the naive design spreads
temperatures out with no regard for the curve's shape.

## Simulating data

Once you have a set of temperatures, `sim_gamma_data()` simulates
performance measurements at them, with observation error drawn from a
gamma distribution (or a truncated-normal approximation when the expected
performance is at or near zero):

```{r sim-data}
dat_opt <- sim_gamma_data(opt_temps, n_reps = 6L, obs_cv = 0.15,
                          ctmin = true_ctmin, ctmax = true_ctmax,
                          a = true_a, b = true_b)
head(dat_opt)
```

## Does the optimized design actually help?

To check whether the optimized temperatures are worth using, we can
simulate data from both designs and fit the Brière-2 curve back to each
dataset, then compare how precisely the true parameters are recovered.
This requires the `nls.multstart` package, so this section is skipped if
it isn't installed:

```{r fit-comparison, eval = requireNamespace("nls.multstart", quietly = TRUE)}
dat_naive <- sim_gamma_data(naive_temps, n_reps = 6L, obs_cv = 0.15,
                            ctmin = true_ctmin, ctmax = true_ctmax,
                            a = true_a, b = true_b)

# Search bounds for starting values, matching how uncertain we said we
# were when calling design_temps() above. These must be given in the same
# order the parameters first appear in the model formula below
# (a, ctmin, ctmax, b).
b_bounds <- exp(log(true_b) + c(-1, 1) * 0.26)
start_lower <- c(a = 0.5, ctmin = true_ctmin - 2, ctmax = true_ctmax - 1.5,
                 b = b_bounds[1])
start_upper <- c(a = 1.5, ctmin = true_ctmin + 2, ctmax = true_ctmax + 1.5,
                 b = b_bounds[2])

fit_tpc <- function(dat) {
    nls.multstart::nls_multstart(
        y ~ a * temp * pmax(temp - ctmin, 0) * pmax(ctmax - temp, 0)^b,
        data = dat, iter = 100,
        start_lower = start_lower, start_upper = start_upper,
        supp_errors = "Y", convergence_count = 30
    )
}

fit_opt <- fit_tpc(dat_opt)
fit_naive <- fit_tpc(dat_naive)

comparison <- rbind(
    truth = c(a = true_a, ctmin = true_ctmin, ctmax = true_ctmax, b = true_b),
    `optimized estimate` = coef(fit_opt),
    `optimized SE` = summary(fit_opt)$coefficients[, "Std. Error"],
    `naive estimate` = coef(fit_naive),
    `naive SE` = summary(fit_naive)$coefficients[, "Std. Error"]
)
round(comparison, 3)
```

In this simulation, the estimates from the optimized design are closer to
the true parameter values, and their standard errors are noticeably
smaller -- especially for `ctmax` and `b`, which the optimized design
samples much more densely near the curve's upper edge. This is the
practical benefit of using `design_temps()`: for the same number of
temperatures and replicates, the resulting fit is more precise.

## Summary

- `briere2_tpc()` and `briere2_tpc_Topt()` let you explore a candidate TPC
  before designing an experiment around it.
- `design_temps()` turns rough prior guesses (and how uncertain you are
  about them) into a set of temperatures expected to be informative for
  fitting that curve.
- `sim_gamma_data()` lets you simulate data from a design before running
  the real experiment, so you can sanity-check it (as we did above) or
  perform a full power analysis.
